---
title: "Assignment 2: Coding Basics"
author: "Laura Brockington"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r, Basics Day 1}
#1. 
seq100x4 <- seq(1,100,4) #creating a sequence from 1 to 100 by 4; sequence is named seq100x4
seq100x4 #printing the output of the sequence code
#2. 
x <- mean(seq100x4) #calculating the mean of the sequence and naming it x
x #printing the mean
y <- median(seq100x4) #calculating the median of the sequence and naming it y
y #printing the median
#3. 
x > y #asking R if the mean is greater than the median
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r, Basics Day 2}
student_names <- c("Jill", "Terry", "Fred", "Kate") #creating a vector of student names
class(student_names) #determining the class of the vector
#student_names is a vector of type character
test_scores <- c(100, 48, 97, 80) #creating a vector of test scores
class(test_scores) #determining the class of the vector
#test_scores is a vector of type numeric
pass_test <- c(TRUE, FALSE, TRUE, TRUE) #creating a vector of whether the student passed the test or not 
class(pass_test) #determining the class of the vector
#pass_test is a vector of type logical

df_student_grades <- as.data.frame(student_names) #creating a data frame with the student names
cbind(df_student_grades, test_scores, pass_test) #adding the test scores and if the student passed the test to the data frame
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame is different from a matrix because it contains multiple types of data (character, numeric, and logical). Matrices can only have one type of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
pass_grade <- function(x) {
  if(x >= 50) {print(TRUE)}
  else {print(FALSE)}
} #creating a function to determine if the student passed the test with if and else
pass_grade2 <- function(x) {
  ifelse(x >= 50, TRUE, FALSE)
} #creating a function to determine if the student passed the test with ifelse

pass_grade2(test_scores) #seeing if the function works on a vector
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Only 'ifelse' worked for my 'test_scores' vector because the if and else function is only designed to work with things that are a length of one. You have to use the ifelse function if you have something with more than one component, such as our vector.
